---
title: "Basic Structure"
author: "Marta Fajlhauer"
date: "15 October 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Installation process: 
Install R: http://www.r-project.org
Install RStudio: http://www.rstudio.com


# Some infos before we will kick off:

This programming workshop wouldn't take place if not a contribution of Dr Jeromy Anglim from Deakin University.
He shared the resources with all members of The R project for Statistical Computing. I used some elements of his workshop here.
This material is used for programming workshop in R for students on the Queen Mary University of London provided by RLadies member (student at QMUL) 
Thank you. 

### What is R and Rstudio??
R is open source programming widely used by statisticians and data miners for developing data analysis.
The name R origin from the fact that two creators of R where Ross Ihaka and Robert Gentleman and that R is build upon statistical programme S. And you know R comes before S, isn't it? 


The most popular interface for R language is RStudio. 


"R is a free software environment for statistical computing and graphics" - http://www.r-project.org


It is a full-featured statistical analysis package (e.g., like SAS, Stata, SPSS, etc.) that allows you to:
Import and export data in a wide range of formats
1. Manipulate data
2. Analyse data
3. Graphically represent data

### Advantages of R:
* It is free to use
* It is open source: it encourages open collaboration. Eveyrone with sufficient skills can create a package to develop furter R. 
* It has a huge library of packages available to run every form of analysis. 
* R is flexible
* R has a nice visualisation graphics that we will see later in the course.

### Challenges of using R
You need to learn programming! ;p It is also fun

#### Introduction to RStudio:

1. File: 
+ Knit document: On more advanced level allows to combine text, R code and Code reuslts into a document like this: 
+ Compile report: Another way of writing a report but I prefer to knit document. 
+ Reopen with encoding: from techopedia: https://www.techopedia.com/definition/948/encoding


Encoding is the process of converting data into a format required for a number of information processing needs, including:
* Program compiling and execution
* Data transmission, storage and compression/decompression
* Application data processing, such as file conversion


Encoding can have two meanings:
* In computer technology, encoding is the process of applying a specific code, such as letters, symbols and numbers, to data for conversion into an equivalent cipher.
* In electronics, encoding refers to analog to digital conversion.


2. Code:

+ Show document ootline: used with R markdown
+ Show diagnostics: Shows different chapters of this document in R. 
+ Extract function: Used in spatial analysis
+ Rename in Scope: This feature makes it easy to rename all instances of a variable
+ Insert Roxygen Skeleton: Showing a full code for that goes behind your panel. For more information go here: https://stackoverflow.com/questions/10825058/can-rstudio-automatically-generate-an-roxygen-template-for-a-function
+ Source with Echo: Source runs the program. Source with Echo runs the program as if you had copied each line into the console, so you see the input followed by the output.


3. Session

+ Setup working directory: All of your work done in R will be saved in the current working directory. 
To setup directory on the beginning of a project fo to file - open new project - ad in pop up window choose directory
To change directory during the project use Session - set working directory

4. Source: It is a place where you can write long codes, projects or documents like this and then run the part of this long document in a console
Check what will happen if you will do 2+2 and Enter
5. Console: It is a place where you run a code and you can see output
6. Environment: all of your saved variables. functions, list, data frames and so on are saved there
7. Files: all files that are avaiable in your working directory
8. Plots: You can see you plots
9. Packages list of all packages downloaded from cran
10. Help
11. Viewer. 


# Packages
R has many additional packages 
To use a package it needs to be installed. 
You only need to install a package once.
To use a package, you need to load the package each time you use R.


# Installation
Option 1. Use the install.packages function.
Note that some packages rely on other packages.
depencies = TRUE ensures that dependencies are also installed.
```{r}
# install.packages("psych", dependencies = TRUE)
```
Option 2. Use the package tab in R Studio
Click install and enter package details

Before we will start anything please dwonload the following packages:
```{r eval=FALSE}
install.packages("knitr", dep = T)
install.packages("MASS", dep = T)
install.packages("psych", dep = T)
install.packages("stringr", dep = T)
install.packages("gdata", dep = T)
install.packages("gtools", dep = T)
install.packages("xtable", dep = T)
install.packages("xlsx", dep = T)
install.packages("readxl", dep = T)
install.packages("lattice", dep = T)
install.packages("ggplot2", dep = T)
install.packages("AER", dep = T)
install.packages("devtools", dep = T)
install.packages("tidyr", dep = T)
install.packages("dplyr", dep = T)
install.packages("Rcmdr", dep = T)
install.packages("lmtest", dep = T)
install.packages("markdown", dep = T)
install.packages("rio", dependencies = T)

library(devtools)
install_github('jeromyanglim/personalityfacets')
install_github('leeper/rio')
```

Loading an installed package
Option 1. Use the library function
```{r}
library(psych) # I.e., put this at the start of your script
```
Option 2.
Click on the package in Rstudio

# Common errors
Not having a package installed is a common error
If you try to load a package that is not installed.
e.g. 
```{r}
library(foo)
```
You will get an error
This means:
1. You mistyped the name of the package, or
2. You need to install the package


# Basic Arithmetic and Logical Operations
```{r basic_multiplication}
# You can use R like a calculator
1 + 5 # addition
10 - 7 # subtraction
10 * 10 # multiplcation
90 / 9 # division
11 ^ 2 # exponentiation

abs(-10) # absolute value
ceiling(2.75) # round up to next integer
floor(3.25) # round down to next integer
sqrt(100) # square roots
exp(2) # exponents
pi # mathematical constant pi
exp(1) # mathematical constant e
log(100) # natural logs (i.e., base e)
log(100, base= 10) # base 10 logs


# Use parentheses to clarify order of operations
(1 + 1 ) * 2
1 + (1 * 2)

# You can test for equality
# TRUE and FALSE are keywords
# T and F are synonyms
TRUE
FALSE
1 == 2 # Equality (Return TRUE if equal)
1 != 2 # Inequality (Return FALSE if unequal)
10 > 9 # Greater than
9 < 10 # Less than
10 <= 10 # Less than or equal
2 %in% c(1, 2 ,3) # is the number in the vector
!(5>1)
result <- 4<5
result2 <- !(5>1) #is not (TRUE)
result | result2 # result 1 or result 2 (because one of them is TRUE)
result & result2 # result 1 and result 2 (because one is not true)
isTRUE(result2)

# TRUE and FALSE coerces to 1 and 0 respectively
as.numeric(TRUE)
as.numeric(FALSE)

# Logical converting to 0, 1 is useful
x <- c(2, 5 ,7 ,10, 15)
x > 5
sum(x > 5)  # sum of a 0-1 variable is a count
mean(x > 5) # mean of a 0-1 variable is a proportion

```

# Basic language features
Assignment:
To assign values to a variable either use "<-" or "="
"<-" is the more common convention R
```{r basic language features, error = TRUE}
x <- 1 + 1
x
```
"=" is the common assignment operator in other programming languages. It does work in R, but is not the convention.
```{r, error=TRUE}
y = 1 + 1
y
```
# Types of variables
```{r Types of variables}
# integer
y <- 2L
typeof(x)
# but double
y <- 2
typeof(y)
# It is because R read by default that you have 2.0 rather than 2. It is exactly why you need to type 2L. 
# complex
z <- 1i
typeof(z)
# but 
z <- i
typeof(z)
# character
a <- "apple"
typeof(a)
# logical
b <- T
typeof(b)
c <- FALSE
typeof(c)
# but
c <- False
typeof(c)
# creating a string
greeting <- "Hello"
name <- "world"
message <- paste(greeting, name)
message
```

### Variable name rules
1. Variable names generally
+ Start with a letter (lower or uppercase)
+ Followed by letters, numbers, underscore (_), or period (.)
+ No spaces

These do not work
```{r, error=TRUE}
my variable <- 1234
1234variable <- 1234
1234 <- 1234
```
This works
```{r, error=TRUE}
myvariable <- 1234
my_variable <- 1234
my_variable <- 1234
myvariable123 <- 1234
myVariable <- 1234
my.variable <- 1234
```
R has many naming conventions
As a matter of preference, style, and convenience, I prefer:
1. Short but descriptive names

* Less than 8 characters for names of lists and data.frames

* Less than 15 characters for variables names in data.frames

2. Use underscore to separate words within a variable name

3. Avoid upper case letters


Names starting with a period are hidden
```{r, error=TRUE}
.myvariable <- 1234
ls()
ls(all.names = TRUE)
```

### Spaces: 
R will generally permits zero, one or more spaces between variables, operators, and other syntactic elements. However, appropriate and consistent spacing improves the readability of you scripts.
[See Hadley Wickham's style guide](http://adv-r.had.co.nz/Style.html) 

This is bad but works
```{r, error=TRUE}
x<-c(1,2,3,400)*2
x<-    c (   1,2,3,      400)*   2
```
This is more readable:
Add spaces after variables, operators, commas
```{r, error=TRUE}
x <- c(1, 2, 3, 400) * 2
```

Multipline line commands: Commands can generally span multiple lines as long as R does not think the command has finished 

This works

```{r, error=TRUE}
x <- c("apple", 
       "banana") 
x
y <- 10 + 
    10  #this works
y

# This does not work
y <- 10
    + 10 
y
```


Multiple commands on one line: You can include more than one command on one line by separat the commands by a semicolon. But generally, you should avoid doing this as it is not very readable.
```{r, error=TRUE}
x <- c(1, 2); y  <- c(3, 4); z <- rnorm(10)
x;y;z
```

R is case sensitive
```{r, error=TRUE}
test <- "lower case"
TEST <- "upper case"
TEST 
test # The original value was not lost
     # because test is different to TEST
Test # This variable does not exist

Test <- "title case"
Test
```

tip: It's often simpler to make variables lower case so that you don't have to think about case.

# Understanding directories
```{r directories}
# R has a working directory.
# This is important when loading and saving files to disk
getwd() # show the current working directory
```
you can use setwd in R studio easily. Go to session -> set working directory
In the Files you can see all folders available in your working directory


Tips: 
* Try to avoid spaces in file names (use hyphen or underscore instead)
* If you do use spaces, then you'll need to escape the space with a slash  (e.g., ("my\ documents")
* Use backslash as the directory separator 
* Store all relevant files for a project within the project working directory

Create some objects in the global environment
```{r}
x <- 1:10
y <- 1:20
data("mtcars")
```

Show objects in the global environment
``` {r}
ls() 
```
or look at the environment panel in RStudio

### Saving objects
Save all objects in the workspace
```{r}
save.image() 

x <- 30
y <- 1:10
```
Save specific named objects using save function.
rdata or RData is the standard file exetnsion.

### Removing objects:
Removing named objects with the rm function
```{r}
rm(x)
ls()
rm(y, mtcars)
```
### Remove all objects from global workspace
Option 1. Use the following command
```{r}
rm(list = ls())
```
Option 2. Click the broom object in RStudio Environment pane

Let's remove x and change y
```{r}
rm(x) 
y <- "changed"
y
```

# Getting help
```{r getting help}
# Use question mark (i.e., ?) followed by command name
# to lookup specific command
?mean 
help(mean) # or use help function

# to look up package
help(package = "MASS")

# Press F1 in RStudio on the command name
# mean 

# Use double question mark to do a full-text search on R help
??"factor analysis"

# Search google
# e.g., how to get the mean of a vector using r

# Ask question on Stackoverflow with the R tag
# http://stackoverflow.com/questions/tagged/r
```

# Importing data
```{r reading data}
# A simple option is to export data from your external data
# in csv format and then import the data using csv
# csv
medals <- read.csv("data/practice/medals.csv")
head(medals)
tail(medals)
dim(medals)

# The read.table() can accept many different types of data separators (commas, tabs, spaces, pipes and others) and it has many options for controlling quoting and escaping data. 
# Can read from local files or remote URLs. 
# If a resource name ends with the .gz suffix, read.table() assumes the file has been compressed in gzip style and will automatically decompress it while reading. 
# Load a data and store it in a new R data frame object called uciCar
uciCar <- read.table(
  'http://win-vector.com/dfiles/car.data.csv',
  sep = ',',
  header = T) # Tell R to expect a header line that defines the data column names
  # colClasses = c('factor', 'factor', rep('integer', 2), rep('factor', 3)) # you can specify what kind of variables you have
  # na.strings = 'NA'
# class() tell what type of object you have. 
class(uciCar)
# summary() gives you a summary of almost any R object
summary(uciCar)
# For data frames, the command dim() show you how many rows and comlumns are in the data
dim(uciCar)
# install the data set from URL source
d <- read.table(
  'http://archive.ics.uci.edu/ml/machine-learning-databases/statlog/german/german.data',
  sep='',
  stringsAsFactors = F,
  header = F)
print(d[1:3,])
# change column names to something meaningful
colnames(d) <- c('Status.of.existing.checking.account',
    'Duration.in.month','Credit.history','Purpose',
    'Credit.amount','Savings.account.bonds',
    'Present.employment.since',
    'Installment.rate.in.percentage.of.disposable.income',
    'Personal.status.and.sex','Other.debtors.guarantors',
    'Present.residence.since','Property','Age.in.years',
    'Other.installment.plans','Housing',
    'Number.of.existing.credits.at.this bank','Job',
    'Number.of.people.being.liable.to.provide.maintenence.for',
    'Telephone','foreign.worker','Good.Loan')
d&Good.Loan <- as.factor(ifelse(d$Good.Loan=1, 'GoodLoan', 'BadLoan'))
print(d[1:3,])

# Other delimited formats
medals <- read.table("data/practice/medals.tsv", sep ="\t")

# Read Excel: 
# Read xls files using xls
# Requires that Perl is installed and on the path
# You may need to install Perl if on Windows
# https://www.perl.org/get.html
# library(gdata) 
# medals <- gdata::read.xls("data/practice/medals.xls")

# requires Java
# library(xlsx)
# x <- xlsx::read.xlsx("data/practice/medals.xlsx", sheetIndex = 1)

# More recent package that has no dependencies on external packages 
# readxl
library(readxl)
medals <- readxl::read_excel("data/practice/medals.xls")
medals <- readxl::read_excel("data/practice/medals.xlsx")


# SPSS
library(foreign)
cas <- foreign::read.spss("data/practice/cas.sav", to.data.frame = TRUE)
attr(cas, "variable.labels")
# tip: You may need to think about value labels in your SPSS file
# Specifically, if you have numeric variables that have variable labels, you may
# want to remove the value labels in SPSS or 

# import stata
?read.dta
?read.sas


# General purpose packages 
# The haven package also can read and write SPSS, SAS, and Stata files

# rio package: General purpose import and export
# General purpose import and export tools
# It's a fairly new package so there may still be a few bugs.
# https://github.com/leeper/rio
# github version is currently a little bit ahead of the CRAN version
library(rio)
cas <- rio::import(file="data/practice/cas.sav")
rio::export(cas, file="output/cas.xlsx")
medals <- rio::import(file="data/practice/medals.csv")

# Use ProjectTemplate to auto-import (see discussion later)
```

# Exporting data 
```{r exporting data}
mydata <- data.frame(a = c(1,2,3), b = c("a", "b", "c"))

# Interal R format
# Good option if you need to re-open data in R
save(mydata, file="output/mydata.rdata")
#  load("output/mydata.rdata")


# csv
# Good option if you need to get data into other software
# This should open in almost all other software (e.g. Excel, SPSS, etc.)
write.csv(mydata, file = "output/mydata.csv")
write.csv(mydata, file = "output/mydata-2.csv", row.names = FALSE) # exclude row.names

# If you need more flexibility in terms of delimiters, etc.
write.table(mydata, file = "output/mydata.tsv", sep = "\t") # e.g., tab delimiter

# Exporting to other formats
# There are a range of options for exporting to other formats
# Functionality is often spread around
# Given that the csv option is usually sufficient
library(foreign)
?foreign::write.foreign  # options for exporting to SAS, SPSS, and Stata directly

?rio::export
```
