---
title: "Different objects in R"
author: "Marta Fajlhauer"
date: "15 October 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Vectors
```{r data types}
# In R, a single value (scalar) is a vector.
x <- 1 # I.e., x is a vector of length 1


# In addition to importing data,
# R has various functions for creating vectors.
c(1, 2, 3, 4) # c stands for combine
1:10 # create an integer sequence 1 to 10
seq(1, 10) # alternative way of creating a sequence 
seq(1, 10, by = 2) # The function has additional options
seg(from=0, to=3, by=0.5) #length not specified. 
seq(from=0, by=0.5, length.out=7) #end not specified
seq(from=0, to=3, lenthg.out=7) #increment not specified
ABC =c("A", "B", "C") #define a vector for replication

rep(1, 5) # repeat a value a certain number of times
rep(c(1,2,3), 5) # repeat a value a certain number of times
rep(ABC, 2)
rep(ABC, times=c(4,2,1))
rep(ABC, each=2)
rep(ABC, each=2, length=10)
rep(ABC, each=2, times=c(1,2,3,1,2,3)) #each=2 argument was applied firstcreating the vector which has six components. adfter that times argument was applied. 
# getting elements of a vector

x = c(2.718, 3.14, 1.414, 47405) #define the vector
z = c(3.45, 2.67)
w = x+z # vector w was created by replicating the z vector 2 times
names(x)= c("e", "pi", "sqrt2", "zipcode") #name the components
x[c(2,4)] #which indices to include
x[c(-1, -3)] #which indices to exclude
x[c("pi", "zipcode")] #names of indices to include
x <- c(FALSE, TRUE) # logical vector
y <- c("a", "b", "cat", "dog") # character vector
z1 <- c(100, 1, 2, 3) # numeric integer vector
z2 <- c(100.2, 0.4, 0.9) # numeric real/double vector
class(x);  typeof(x); mode(x)
class(y);  typeof(y); mode(y)
class(z1);  typeof(z1); mode(z1)
class(z2);  typeof(z2); mode(z2)

# Checking type of object
is.logical(c(TRUE, TRUE))
is.numeric(c("a", "b"))
is.character(c(1, 2, 3))

# Conversion of Types:
as.character(c(1, 2, 3, 4))
as.numeric(c("1", "2a", "3", "four"))
as.numeric(c(FALSE, FALSE, TRUE, TRUE))

# R often performs conversions implicitly
sum(c(FALSE, TRUE, TRUE)) # converts logical to 0, 1 numeric
paste0("v", c(1, 2, 3)) # converts numeric vector to character
```

### Sample:
```{r basic data structures}

# as well as many simulation functions which we'll cover later
# Initial examples:
# Sample 10 items with replacement from 
sample(x = c("happy", "funny", "silly"), size = 10, replace = TRUE)

# Sample 20 values from a normal distribution
y <- rnorm(n = 20, mean = 0, sd = 1) 
hist(y) # show values in histogram


# Vectors can have names
x <- c(1,2,3,4,5)
names(x) <- c("a", "b", "c", "d", "e")
x

# Extracting vectors
x[c(1,2)] # by numeric position
x[x < 3] # by logical vector
x[c("b", "c")] # by name
```

### Factors
```{r}
# Factors are a type of vector in R for which the elements are categorical values that could also be ordered. The values are stored internally as integers with labeled levels. 
x= c("high", "medium", "low", "high", "medium")
xf = factor(x) #Notice that R extracted "levels" from the vector and listed them after the contents of the factor. By defaUlt it ordered the levels alphbetically. 
as.numeric (xf) #show factors as numeric vectors
xfo = factor(x, levels=c("low", "medium", "high"), ordered = TRUE) #specify a particular order for the levels
as.numeric(xfo)
# reordering the levels of a factor that has already been created. 
x= c("high", "medium", "low", "high", "medium")
xf = factor(x)
xf 
# Now we reorder the levels by applying the factor function to the existing factor xf:
xf = factor(xf, level =c("low", "medium", "high"), ordered = TRUE)
xf
# But for resulting output, the levels can be relabeled to whatever might be more meaningful for the application, e.g.
xfol = factor (x, levels=c("low", "medium", "high"), ordered = TRUE,
               labels=c("Bottom SES", "Middle SES", "Top SES"))
xfol
# it is important to know that that relabeling the levels does not change their order or their integer coding.
```
### Matrices:
All data must be of same type (e.g., numeric, character, logical)
```{r}
# A matirx is a two-dimensional array of values of the same type. 
matrix(1:6, ncol=3) # contents are 1:6, filled by column. 
matrix(1:6, nrow=2) # or you can specify number of rows. 
matrix(1:6, nrow=2, byrow=TRUE) # filled by row instead of by column 
x = matrix(1:6, nrow=2, 
       dimnames=list(TheRowDimName = c("Row1Name", "Row2Name"),
                     TheColDimName = c("Col1Name", "Col2Name", "Col3Name")))
# The components of matrices can be accessed via their indices or their names. The first index refers to the row and the second refer to the columns. 
x[2,3]
x["Row2Name", "Col3Name"]
x[2, 1:3] # specify a range of columns
x[2,]
x[,3]
y <- matrix(c(1, 2, 
             4, 5, 
             7, 8 ),
           byrow = TRUE, ncol = 2)
y

class(y)

dim(y)  # Number of rows and columns
nrow(y) # Number of rows
ncol(y) # Number of columns

# Rows and columns can be given names
rownames(y) <- c("a", "b", "c")
colnames(y) <- c("col1", "col2")

# Rows and columns can be indexed
y["a", ] # By rowname
y[, "col1"] # By column name 
y["a", "col1"] # By both

y[c(1,2), ] # By row position
y[,1] # By column position
y[c(2,3), 2] # By column position
```

### Lists
* Store arbitrary structures of one or more named elements.
* Elements can be of different lengths
* Lists can contain lists can be nested to create tree like structures
* Lists are commonly used for representing results of analyses
```{r)}
MyList = list("a"=1:3, b=matrix(1:6, nrow=2), "c"="Hello, world")
MyList$a #the content of the first list item named "a"
MyList[[1]]
MyList$a[2] #the second element of the list item named "a"
MyList[[1]][2]
w <- list(apple = c("a", "b", "c"), 
          banana = c(1,2), 
          carrot = FALSE,
          animals = list(dog = c("dog1", "dog2"), 
                         cat = c(TRUE, FALSE)))

class(w)

# Accessing one element of list
w$apple # using dollar notation
w[[1]] # by position
w[["apple"]] # by name (double brackets)

# Accessing subset of list
w[c(1, 2)]  # by position (single bracket)
w[c("apple", "banana")] # by name
w[c(FALSE, FALSE, TRUE, TRUE)] # by logical vector

# Quick illustration of a list object returned by
# a statistical function

# We'll simulate some data for two hypothetical groups x and y
# and perform an independent samples t-test.
x <- rnorm(10, mean = 0, sd = 1) 
y <- rnorm(10, mean = 1, sd = 1)
fit <- t.test(x, y)  

# The function 
class(fit) # class does not say list, but it is a list
mode(fit)
str(fit) # show structure of object
names(fit) # show names of elements
# we can view particular elements
fit$statistic 
fit$parameter
fit$p.value
# or extract subsets of the list
fit[c("statistic", "parameter", "p.value")]
```
### Data Frames:
* Data frames are the standard data strucure used for storing data. If you have used other software (e.g., SPSS, Excel, etc.), this is what you may think of as a "dataset".
* Columns can be of different data types (e.g., character, numeric, logical, etc.)
```{r}
z <- data.frame(var1 = 1:9, var2 = letters[1:9])
z
```
### Tip: 
* Some functions work with matrices, some work with data.frames, and some work with both.
* If you are wanting to store data like you might store in a database, then you'll generaly want a data.frame.
* If you are dealing with a mathematical object that you you want to perform a mathematical operation on, then you generally want a matrix (e.g., correlation matrix, covariance matrix, distance matrix in MDS, matrices used for matrix algebra).

# Working with data frames
```{r exploring data frames}
# Let's use the built-in survey data.frame dataset
library(MASS)
data(survey)
?survey
mydata <- na.omit(survey) # for simplicity I'll exclude missing data
shortdata <- mydata[1:6, 1:5] 
shortdata
```
### Extracting observations (i.e., rows) and variables (i.e., columns).
There are similarities to matrices and lists
Select observations
```{r}
shortdata[1:5, ] # by row number
shortdata[c(5,4,3,2,1), ] # re-order
shortdata[ shortdata$Sex == "Female", ] # by logical vector
shortdata[c("1", "2"), ] # by rownames

# Select variables 
shortdata[, c(1,2)] # by position like a matrix
shortdata[c(1,2)] # by position like a list
shortdata[ ,c("Sex", "Fold")] # by name like a matrix
shortdata[c("Sex", "Fold")] #
shortdata$Sex # by name to get a single variable
```

### Names
```{r}
names(shortdata) # get variable names
colnames(shortdata) # but this also works
rownames(shortdata) # rows can also have names
```
#### Tip: 
Avoid row names. Add another variable to the data.frame to store this information.


### Examine first few rows
```{r}
head(mydata) # first 6 rows
head(mydata, n = 10) # first 7 rows
tail(mydata) # last few rows
View(mydata) # Rstudio function to open data in viewer or click on the icon in the Environment pane

# How many rows and columns?
dim(mydata) # rows and column counts
nrow(mydata) # row count
ncol(mydata) # column count

# Examine structure
str(mydata)
```

### Functions
# A function takes input values, called "arguments" and does something with them. In general a function in R is defined by code of the form:
# functionName = function(arguments){commands}
# When the function is called it takes the values of the arguments in parentheses and use them in the commands in the braces
``` {r}
asqplusb = function(a, b=1){
  c=a^2+b
  return(c)
}
asqplusb(a=3, b=2)
# However arguments without explicit labels must be in the order used in the definition of the function
asqplusb(3,2)
asqplusb(2,3)
```
# The function definition gave argument b a default value of 1 by specifying b=1 in the list of arguments. This means that if the function is called 
# without an explicit value for b provided, the default value will be used. The argument a was not given the default value, however, and must be always specified 
# in a call to the function.
# String functions
```{r string functions}
paste("hello", "how", "are", "You") # defaults to space separator
paste0("hello", "how", "are", "You") # no separator
paste("apple", "banana", "carrot", "date", sep =", ") # specify arbitrary separator
paste0("v", 1:10) # paticularly useful with vectors

# Extract substring
substr("abcdefghijklmnop", 4, 6)

# Change case
toupper("abcd") # make upper case
tolower("ABCD") # make lower case

mystring <- c("apple", "banana", "carrot", "date", "egg", "fig")
# Identify which strings match a pattern
grep("a", mystring) # index of objects with "a"
grep("a", mystring, value = TRUE) # value of objects with "a"

# get count of number of characters
nchar(mystring)
data.frame(mystring, nchar(mystring))

# Substitute a mystringreplacement text that matches a pattern
questions <- c("How are you?", "What is going on?")
gsub(" ", "_", questions) # replace space with underscore


# R  string manipulation tools are very powerful
# For more information see
?grep
?"regular expression"


# see also Hadley Wickham's package for string manipulation
# It attempts to introduce greater consistency in notation.
# install.packages("stringr")
library(stringr)
help(package = "stringr")
# all functions begin with str_
str_length(mystring) # see nchar
str_sub(mystring, start = 1, end = 3)


# writing output to the console
cat("Hello World!")

# Tab is \t and new line is \n
cat("Hello\t World\nSome more text")

```

### Loops

We very raly use loops in R as in R much quicker is to use different packages or functions working in similar way as you want the loop to work in
We will cover different types of loops but please bear in mind that structure of loop is important
### while (logical expression){body of the loop}
R will check the logical expression then R will execute whatever is in the body of the loop. It will continue until the logical expression is false. Then it will stop
```{r while loop}
while(FALSE){
  print('Hello')
} 
```
we shouldn't see anything as the logical statement is not TRUE
while(TRUE){
print('Hello')
} This loop would continue forever till we want press the red stop button that we would see in  console
Let see an example:
```{r}
counter <- 1
while(counter < 12){
  print(counter)
  counter <- counter + 1
}
```
#### for loop
for (specify iteration for the loop){body of the loop}
```{r}
for (i in 1:5){
  print('Hello R')
}
```
### if loop
if (logical expression but run only once){body of the loop}
```{r}
x <- rnorm(n = 1) # generate 1 random number from normal distribution with mean=0 and st.dev=0
if(x >1){
  answer <- 'Greater than 1'
} # if answer variable not greater than x !> 1

rm(x)
x <- rnorm(n = 1)
if(x >1){
  answer <- print('Greater than 1')
} else {
  answer <- print('Less than 1')
}# one or the other is executed

rm(x)
x <- rnorm(n = 1)
if(x >1){
  answer <- print('Greater than 1')
} else {
  if(x >= -1){
    answer <- print('between -1 and 1')
  } else{
    answer <- print('less than -1')
  }
}
# Below you have exactly the same loop as above only a little bit less confusing
rm(answer, x)
x <- rnorm(n=1)
if(x > 1){
  answer <- 'Greater than 1'
} else if (x >= -1){
  answer <- 'between -1 and 1'
} else{
  answer <- 'Less than 1'
}
```
